The role of aneuploidy in the evolution of cancer drug resistance

### 2023-06-19 EvolutionaryRescue1.tex
added parameter values to figures

In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
\begin{align*}
\lambda_w' &= \lambda_w, \\
\mu_w' &= \mu_w,\\
\lambda_a' &= C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\\ 
\mu_a' &= C_1,\\
\lambda_m' &= C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\\ 
\mu_m' &= C_2,
\end{align*}
where $C_1, C_2>0$ are constants. % TODO what constants? see Remus version.

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
    \hline
     & Name & Value & Units & References \\ \hline
    $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\lambda_a$  & Aneuploid division rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ \hline
  \end{tabular}
\caption{\textbf{Model parameters.} 
Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
  \label{table1}
\end{center}
\end{table}

@@

\begin{figure}
% TODO change panel labels to ABCD
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
\end{subfigure}
\caption{
\textbf{Evolutionary rescue probability with partially resistant or tolerant aneuploid cells.}
Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{Aleft}, $N=10^4$) and even more so in an initially large tumor (\textbf{Aright}, $N=10^8$). 
When aneuploidy provides tolerance (\textbf{Bleft}, $N=10^4$; \textbf{Bright}, $N=10^8$), the rescue probability is much lower. 
In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
Here, division rate $\lambda_w=\lambda_a=\lambda_m=0.14$, (so that growth rate changes due to variable death rate), mutant death rate $\mu_m=0.13$ (so that mutant growth rate $\Delta_m=0.01$), aneuploidy rate $u=10^{-2}$, mutation rate $v=10^{-7}$.
}
\label{rescue_prob_wt_growth}
\end{figure}
%%%%%%%%%%%

\begin{figure}
% small population size
% switch from tolerance to non-growing
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf(A) $N=10^4$; \textbf{(B)} $N=10^8$). 
Markers for simulation results with 95\% CI; 
solid lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob});
dashed lines for the approximate formula (\cref{rescue_prob_approx}).
The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}$.
% TODO why no switch in panel B?
% TODO add parameter values
% TODO more simulations for 0 > Delta_a > Delta^*_a
}
\label{rescue_prob_an_growth}
\end{figure}

@@

\begin{figure}
\begin{subfigure}{0.5\textwidth}
% non-growing slightly declining, compare 1st order approximation to exact and simulation
\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
% density-dependent simulations, barely growing (or maybe partial resistance), with exact formula
\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue probability for variable initial tumor size.}
\textbf{(A)} Comparison of simulation results (markers with 95\% CI, too small to appear with $10^5$ simulations per marker), the exact formula (blue line, \cref{survival_prob} in \cref{rescue_prob}) and the approximate formula (orange line, \cref{rescue_prob_approx}).
\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{survival_prob} in \cref{rescue_prob}). Here, $K=10^9$.
% TODO add parameter values
}
\label{rescue_prob_N}
\end{figure}

@@

\begin{figure}
% slightly decreasing (need to check)
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
\caption{\textbf{Effect of standing variation on evolutionary rescue.}
In aneuploid cells already exist in the population at the onset of drug therapy as standing genetic variation, then evolutionary rescue is more likely... 
% TODO think about this figure...
% TODO maybe better to show p_total / p_denovo as function of u or N.
% TODO add parameter values
}
\label{FractionPlot}
\end{figure}

@@

\begin{figure}
% A partial resistance
% B tolerance
% left function of wildtype growth rate 
% right function of population size
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
\end{subfigure}
\caption{\textbf{Effect of aneuploidy on evolutionary rescue.}
The ratio of rescue probability with and without aneuploid ($H$, \cref{ratiorescue}) increases with the aneuploid growth rate (colors) and decreases with the wildtype growth rates and initial tumor size (x-axes), except for large tumors where where the ratio converges to unity.
\textbf{(A-left,A-right)} Aneuploidy provides partial resistance.
\textbf{(B-left, B-right)} Aneuploidy provides tolerance.  
Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
% TODO parameter values
% TODO start y at 1
% TODO fix panel labels
}
\label{rescue_ratio}
\end{figure}

@@

\begin{figure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue time.}
Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue \textbf{(left)} with ($u>0$) and \textbf{(right)} without ($u=0$ aneuploidy.
Our inhomogeneous Poisson-process approximations (solid blue lines, right: \cref{meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (orange markers with 95\% CI).
Our 1st-order (dashed red lines, right: \cref{limitapprox}, left: \cref{limitapprox2}) and 2nd-order (green line, left: \cref{limitapprox3}) approximations work well when the initial tumor size is large (here $>10^8$ cells).
% TODO parameter values
}
\label{MeanTimeGrowthAneuploidyPlot} 
\end{figure}

@@

\begin{figure}
% diffusion approx instead of branching process
% tolerance
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the Feller diffusion approximation \eqref{FellerApprox}. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/w_0}$ where $p$ is the mean probability of evolutionary rescue.}
\label{P_est_diffusion}
\end{figure}

\begin{figure}
% partial resistance, compare 1st and 2nd order approximation and exact
\vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells. The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations. For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.135,\mu_m=0.13$. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
\label{compare_1st_2nd_order_approx}
\end{figure}

>>>>

In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
\begin{align*}
\lambda_w' &= \lambda_w, \\
\mu_w' &= \mu_w,\\
\lambda_a' &= C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\\ 
\mu_a' &= C_1,\\
\lambda_m' &= C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\\ 
\mu_m' &= C_2,
\end{align*}
where $C_1, C_2>0$ are constants and $K$ is the maximum carrying capcity. % TODO what constants? see Remus version.

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
    \hline
     & Name & Value & Units & References \\ \hline
    $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\lambda_a$  & Aneuploid division rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ 
    % TODO citing Nowak 2004 for the rates is strange because it is a theory paper - cite the original empirical papers
    \hline
  \end{tabular}
\caption{\textbf{Model parameters.} 
Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
  \label{table1}
\end{center}
\end{table}

@@

\begin{figure}
% TODO change panel labels to ABCD
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
\end{subfigure}
\caption{
\textbf{Evolutionary rescue probability with partially resistant or tolerant aneuploid cells.}
Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{Aleft}, $N=10^4$) and even more so in an initially large tumor (\textbf{Aright}, $N=10^8$). 
When aneuploidy provides tolerance (\textbf{Bleft}, $N=10^4$; \textbf{Bright}, $N=10^8$), the rescue probability is much lower. 
In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
Parameters: division rate $\lambda_w=\lambda_a=\lambda_m=0.14$ (so that growth rate changes due to variable death rate); mutant death rate $\mu_m=0.13$ (so that mutant growth rate $\Delta_m=0.01$); aneuploidy rate $u=10^{-2}$; mutation rate $v=10^{-7}$.
}
\label{rescue_prob_wt_growth}
\end{figure}
%%%%%%%%%%%

\begin{figure}
% small population size
% switch from tolerance to non-growing
% TODO why no switch in panel B?
% TODO more simulations for 0 > Delta_a > Delta^*_a
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf(A) $N=10^4$; \textbf{(B)} $N=10^8$). 
Markers for simulation results with 95\% CI; 
solid lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob});
dashed lines for the approximate formula (\cref{rescue_prob_approx}).
The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}$.
Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
}
\label{rescue_prob_an_growth}
\end{figure}

@@

\begin{figure}
\begin{subfigure}{0.5\textwidth}
% non-growing slightly declining, compare 1st order approximation to exact and simulation
\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
% density-dependent simulations, barely growing (or maybe partial resistance), with exact formula
\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue probability for variable initial tumor size.}
\textbf{(A)} Comparison of simulation results (markers with 95\% CI, too small to appear with $10^5$ simulations per marker), the exact formula (blue line, \cref{survival_prob} in \cref{rescue_prob}) and the approximate formula (orange line, \cref{rescue_prob_approx}).
\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{survival_prob} in \cref{rescue_prob}) with maximum carrying capacity $K=10^9$.
Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.15$, (B) $\mu_a=0.135$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
}
\label{rescue_prob_N}
\end{figure}

@@

\begin{figure}
% slightly decreasing (need to check)
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
\caption{\textbf{Effect of standing variation on evolutionary rescue.}
In aneuploid cells already exist in the population at the onset of drug therapy as standing genetic variation, then evolutionary rescue is more likely... 
Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
% TODO think about this figure...
% TODO maybe better to show p_total / p_denovo as function of u or N.
}
\label{FractionPlot}
\end{figure}

@@

\begin{figure}
% A partial resistance
% B tolerance
% left function of wildtype growth rate 
% right function of population size
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
\end{subfigure}
\caption{\textbf{Effect of aneuploidy on evolutionary rescue.}
The ratio of rescue probability with and without aneuploid ($H$, \cref{ratiorescue}) increases with the aneuploid growth rate (colors) and decreases with the wildtype growth rates and initial tumor size (x-axes), except for large tumors where where the ratio converges to unity.
\textbf{(A-left,A-right)} Aneuploidy provides partial resistance.
\textbf{(B-left, B-right)} Aneuploidy provides tolerance.  
Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
Parameters: $N=10^4$; $\lambda_w=\lambda_a=\lambda_m=0.14$; (B) $\mu_w=0.17$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
% TODO start y at 1
% TODO fix panel labels
}
\label{rescue_ratio}
\end{figure}

@@

\begin{figure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue time.}
Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue \textbf{(left)} with ($u>0$) and \textbf{(right)} without ($u=0$ aneuploidy.
Our inhomogeneous Poisson-process approximations (solid blue lines, right: \cref{meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (orange markers with 95\% CI).
Our 1st-order (dashed red lines, right: \cref{limitapprox}, left: \cref{limitapprox2}) and 2nd-order (green line, left: \cref{limitapprox3}) approximations work well when the initial tumor size is large (here $>10^8$ cells).
Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
}
\label{MeanTimeGrowthAneuploidyPlot} 
\end{figure}

@@

%%%%%%%%%%
%
%\begin{figure}
%% diffusion approx instead of branching process
%% tolerance
% \vspace*{1\baselineskip}
%\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
%\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the Feller diffusion approximation \eqref{FellerApprox}. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/w_0}$ where $p$ is the mean probability of evolutionary rescue.}
%\label{P_est_diffusion}
%\end{figure}
%
%%%%%%%%%%%
%
%\begin{figure}
%% partial resistance, compare 1st and 2nd order approximation and exact
%\vspace*{1\baselineskip}
%\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
%\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells. The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations. For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.135,\mu_m=0.13$. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
%\label{compare_1st_2nd_order_approx}
%\end{figure}
